---
title: "Introduction to multihostSIR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to multihostSIR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 96
)
```

This vignette is the quickest way in. It builds one community, runs one
epidemic, then scales up to the replicate study that is the package's main job.
For the fuller treatment of the model and its assumptions, see
`vignette("study", package = "multihostSIR")`.

## Why multi-host models

Many wildlife and zoonotic pathogens circulate in several host species at once:
SARS-CoV-2, highly pathogenic avian influenza, rabies, bovine tuberculosis. A
single-host SIR model cannot represent the three features that usually dominate
such systems:

- transmission that is **asymmetric** between species,
- contact structure imposed by the **food web**,
- species-specific **susceptibility and virulence**.

`multihostSIR` handles all three, and derives parameters from body mass whenever
empirical values are unavailable.

```{r library}
library(multihostSIR)
set.seed(42)
```

## The three building blocks

Every simulation is assembled from three objects. Each has sensible defaults and
each prints itself, so you can inspect what a run will do before starting it.

```{r blocks}
allo   <- allometry()
spec   <- community_spec()
coeffs <- trophic_coefficients()

allo
```

`allometry()` holds the body-mass power laws `Y = a * M^b`, with `M` in kg and
time measured in days. `community_spec()` controls how a random community is
drawn (the fraction of prey, the body-mass pools, how far predator densities are
thinned). `trophic_coefficients()` sets the strength of each transmission route
between trophic groups.

The allometric relationships are worth seeing directly:

```{r allometry-plot}
plot_allometry(allo)
```

## One community

```{r community}
comm <- build_community(6, spec = spec, allo = allo)
comm[, c("species", "type", "mass", "mu", "gamma", "N", "competence")]
```

Each species gets a trophic type, a body mass drawn from the matching pool,
allometrically scaled rates, and a within-species transmission rate
`beta_ii = competence / N`.

## Transmission between species

```{r transmission}
beta <- transmission_matrix(comm)
round(beta, 6)
```

Diagonal entries are the within-species rates. Off-diagonal entries are the
geometric mean `sqrt(beta_ii * beta_jj)` scaled by a trophic coefficient. Rows
receive infection; columns are the source. Under the defaults, predators pick up
infection from prey but never pass it back, which shows up as a block of zeros:

```{r blocked}
beta[comm$type == "Prey", comm$type == "Predator"]
```

```{r heatmap, fig.height = 5}
plot_transmission_matrix(comm)
```

## Community R0

The community basic reproduction number is the dominant eigenvalue of the
next-generation matrix.

```{r r0}
community_R0(comm, beta)
```

## One epidemic

```{r epidemic}
sim <- simulate_epidemic(comm, beta = beta, duration = 365)
species_metrics(sim)
```

The epidemic is seeded in the most competent host by default. `species_metrics()`
reports, per species, the peak number infected, the day it peaked, cumulative
infections, and the attack rate.

## Replicate study

`run_multihost_sir()` is the entry point you will use most. It draws many random
communities at each richness level, runs an epidemic in each, and records the
outcomes along with per-level timing. Iteration counts are kept small here so the
vignette builds quickly; use hundreds or thousands for real work.

```{r run}
res <- run_multihost_sir(
  richness_levels = 3:6,
  n_iterations    = 15,
  duration        = 180,
  seed            = 42
)

res
summary(res)
```

```{r figures}
plot_r0_richness(res)
plot_trajectories(res, xlim = c(0, 120))
```

## Using your own community

The run above generates communities at random. If you already know your system,
you can supply it directly. A complete species table -- every rate for every
species -- combined with a transmission matrix leaves nothing to randomise, so
the model runs once and returns an exact answer.

```{r fixed}
species <- data.frame(
  species    = c("mouse", "deer", "fox"),
  type       = c("Prey", "Prey", "Predator"),
  mass       = c(0.02, 50, 5),
  mu         = c(0.013, 0.002, 0.003),
  gamma      = c(0.27, 0.04, 0.07),
  alpha      = c(0.027, 0.004, 0.007),
  N          = c(9000, 27, 6),
  competence = c(0.5, 0.1, 0.2)
)

comm <- read_community(species)
attr(comm, "complete")   # TRUE: nothing needed filling in

beta <- read_transmission_matrix(matrix(c(
  0.05, 0.01, 0.00,
  0.01, 0.03, 0.00,
  0.02, 0.02, 0.04), nrow = 3, byrow = TRUE), community = comm)

fit <- run_fixed_community(comm, beta, duration = 200)
fit
```

`read_community()` also accepts a path to a `.csv` or `.xlsx` file. Any column
you leave out is filled from body mass, and the printed "assumptions used" block
records what was supplied versus filled. If you give a community but no matrix,
the matrix is built from your competence values and the trophic rules; if your
table is only partial, the gaps are generated and you are back to averaging over
`run_multihost_sir()`.

## Where to next

- `vignette("study", package = "multihostSIR")` for the model definition,
  run-time scaling, reproducibility, and how to change every assumption.
- `?run_multihost_sir` and the other help pages, each with runnable examples.
- `demo(getting_started, package = "multihostSIR")` for a console walkthrough.
