---
title: "The multi-host SIR model and simulation study"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{The multi-host SIR model and simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 96
)
library(multihostSIR)
set.seed(42)
```

This vignette covers the model itself, the scaling assumptions, run-time
behaviour, reproducibility, and how to change the defaults. If you have not
already, read `vignette("multihostSIR")` first for the basic workflow.

## Model definition

For species $i$ in a community of $n$ species:

$$\frac{dS_i}{dt} = b_i - \mu_i S_i - \Lambda_i S_i$$
$$\frac{dI_i}{dt} = \Lambda_i S_i - (\mu_i + \gamma_i + \alpha_i) I_i$$
$$\frac{dR_i}{dt} = \gamma_i I_i - \mu_i R_i$$
$$\frac{dC_i}{dt} = \Lambda_i S_i$$

where $\Lambda_i = \sum_j \beta_{ij} I_j$ is the force of infection on species
$i$, and $C_i$ accumulates cumulative incidence without feeding back into the
dynamics.

Births $b_i = \mu_i N_i$ are held constant, so the disease-free equilibrium is
$S_i = N_i$. Total population size is **not** conserved once virulence
$\alpha_i > 0$. This is a modelling choice, not an oversight: it holds the
disease-free state at the allometric carrying capacity while letting
disease-induced mortality draw down standing abundance.

A discrete event fires at each integer time step and sets any infected
compartment holding fewer than one individual to exactly zero. Without it,
floating-point residues on the order of $10^{-15}$ individuals sustain phantom
transmission indefinitely. Disable it with `extinction_threshold = 0`.

## Allometric scaling

If a species' rates are not measured, they are derived from body mass $M$ (kg)
via $Y = a M^{b}$. The unit of time is one day.

| Symbol   | Meaning                              | Default            |
|----------|--------------------------------------|--------------------|
| $\mu$    | background mortality (per day)       | $0.005\,M^{-0.25}$ |
| $\gamma$ | recovery rate (per day)              | $0.1\,M^{-0.25}$   |
| $N$      | carrying-capacity density (per km^2) | $500\,M^{-0.75}$   |
| $\alpha$ | disease-induced mortality (per day)  | $0.01\,M^{-0.25}$  |

The exponents are not arbitrary. Physiological rates and lifespan scale as
roughly $M^{-0.25}$, so larger animals live longer and recover more slowly
(Kleiber 1932; Lindstedt and Calder 1981; Peters 1983). Population density
scales as $M^{-0.75}$ by Damuth's law, so larger animals are rarer
(Damuth 1981).

```{r allometry}
allo <- allometry()
allo

# a small host against a large one
rbind(small = allo$a_mort * 0.02^allo$b_mort,
      large = allo$a_mort * 50^allo$b_mort)
```

## Transmission and R0

The between-species transmission matrix has diagonal
$\beta_{ii} = \mathrm{competence}_i / N_i$ and off-diagonal entries
$c \cdot \sqrt{\beta_{ii}\beta_{jj}}$, where $c$ is a trophic coefficient. The
next-generation matrix is

$$K_{ij} = \beta_{ij}\, N_i\, D_j, \qquad D_j = \frac{1}{\gamma_j + \mu_j + \alpha_j}$$

and community R0 is its dominant eigenvalue.

```{r r0}
comm <- build_community(6)
beta <- transmission_matrix(comm)
community_R0(comm, beta)
```

One subtlety: R0 is *invariant* when every species' density is scaled by the same
factor, because $\beta_{ii} = \mathrm{competence}/N$ falls as $1/N$ while $K$
carries a factor of $N$, and the two cancel. Density *differences* between species
still matter; a uniform rescaling does not.

```{r r0-invariance}
comm2 <- comm
comm2$N <- comm2$N * 10
comm2$beta_ii <- comm2$competence / comm2$N
c(original = community_R0(comm), rescaled = community_R0(comm2))
```

## Run-time scaling

Each richness level is timed separately, which turns a replicate run into a
scaling study.

```{r timing-run}
res <- run_multihost_sir(
  richness_levels = 3:7,
  n_iterations    = 10,
  duration        = 150,
  seed            = 42
)

cx <- timing_complexity(res)
c(exponent = cx$exponent_richness, r_squared = cx$r_squared_richness)
```

```{r timing-plot}
plot_timing(res, log_scale = TRUE)
```

The $n \times n$ matrix-vector product in the force-of-infection term and the
$n \times n$ eigendecomposition used for R0 make an exponent between 2 and 3 the
expectation at large $n$. A short untimed warm-up batch runs before the timed
loop so that one-off costs (JIT compilation, cluster start-up) do not inflate
the first level; disable it with `warmup = FALSE`.

## Reproducibility

Same seed, same results:

```{r repro}
a <- run_multihost_sir(3:5, n_iterations = 10, duration = 100, seed = 42)
b <- run_multihost_sir(3:5, n_iterations = 10, duration = 100, seed = 42)
identical(a$results$R0, b$results$R0)
```

Serial runs use `set.seed()`; parallel runs use
`parallel::clusterSetRNGStream()`, applied after the warm-up so warm-up never
shifts the recorded results. A serial run and a parallel run with the same seed
give different (but equally valid) results, because L'Ecuyer streams differ from
the default RNG. Fix `parallel`, `n_cores`, and `seed` together for bit-identical
output.

## Changing the assumptions

Every default is an argument. Here predators are less rare, prey are larger,
recovery is slower, and the normally blocked predator-to-prey route is opened.

```{r custom}
custom <- run_multihost_sir(
  richness_levels = 3:5,
  n_iterations    = 10,
  duration        = 150,
  spec = community_spec(
    predator_density_factor = 0.4,
    prey_masses = c(5, 20, 80)
  ),
  allo = allometry(a_recovery = 0.05),
  coefficients = trophic_coefficients(pred_to_prey = 0.1),
  seed = 1
)

summary(custom)
```

## Exporting results

```{r export, eval = FALSE}
export_results(res, dir = "output", formats = c("csv", "xlsx"))
```

This writes the results, trajectories, timing, summary, and derived tables.
Excel output needs the optional `openxlsx` package; without it you get a warning
and CSV only. Excel sheets cap at 1,048,576 rows, so larger tables get a pointer
note in the workbook while the complete data always go to CSV.

## Larger runs

For production work, raise the iteration count and go parallel:

```{r big, eval = FALSE}
big <- run_multihost_sir(
  richness_levels = 3:30,
  n_iterations    = 10000,
  seed            = 42,
  parallel        = TRUE,
  n_cores         = parallel::detectCores() - 1,
  progress        = TRUE
)
```

`parallel` defaults to `FALSE` and `n_cores` to `2`, because package checks must
not use more than two cores. Raise both on your own machine.

## References

Damuth, J. (1981) Population density and body size in mammals. *Nature* **290**,
699-700.

Diekmann, O., Heesterbeek, J.A.P. and Metz, J.A.J. (1990) On the definition and
the computation of the basic reproduction ratio R0 in models for infectious
diseases in heterogeneous populations. *Journal of Mathematical Biology* **28**,
365-382.

Kleiber, M. (1932) Body size and metabolism. *Hilgardia* **6**, 315-353.

Lindstedt, S.L. and Calder, W.A. (1981) Body size, physiological time, and
longevity of homeothermic animals. *Quarterly Review of Biology* **56**, 1-16.

Peters, R.H. (1983) *The Ecological Implications of Body Size*. Cambridge
University Press.
